Large language models (LLMs) impose significant computational and financial costs through token-based pricing, with costs scaling linearly with input prompt length. We present a model-free statistical filtering approach that achieves 50\% token reduction while maintaining 91\% quality retention, validated across 6 flagship LLMs with 350+ A/B test pairs. Unlike existing methods that rely on external language models (e.g., LLMLingua), our approach uses pure statistical heuristics—inverse document frequency (IDF), position-based importance, part-of-speech filtering, and named entity detection—to identify and remove low-value tokens. The algorithm operates in $O(n \log n)$ time where $n$ is word count, achieving 10.58 MB/s throughput on modern hardware. Real-world validation on Grok-4 (93\% quality), Claude 3.5 Sonnet (91\% quality), GPT-5 (89\% quality), and Gemini Pro (89\% quality) confirms production readiness with 92\% keyword retention and 89.5\% entity retention. Our implementation in Rust provides a production-ready solution with comprehensive quality metrics and configurable compression levels (30\%, 50\%, 70\%). Additionally, we introduce optical context compression (BETA), rendering compressed text as 1024×1024 images for vision model consumption. For Claude Sonnet pricing (\$15/1M input tokens), this translates to \$7.50 saved per million tokens, or \$900K/year for enterprise applications processing 1B tokens monthly. The algorithm is transparent, deterministic, and requires no external dependencies, making it suitable for offline deployment and cost-sensitive production systems.
